The late 1800's debate over the age of the Earth: an early attempt of interdisciplinary science

### 2023-08-17 ms.tex
wordsmith

Many of the big unsolved problems in science today will require an interdisciplinary approach to answer. For example, the unsolved problem of the origin of life on Earth can only be tackled using many different fields of research \citep{Hays_2015}. Geologic expertise is required to understand the earliest fossil records of life, organic chemists are needed to synthesize life-like replicating molecules, and atmospheric modelers are essential to estimate the plausible conditions on the early Earth under which life began. Another problem that requires broad expertise is projecting the impact of anthropogenic climate change on society. Any forecast will involve atmospheric scientists, glaciologists, economists, and ecologists among many other areas of knowledge. Given that important unanswered problems rely on an interdisciplinary approach, it is essential that scientists work together effectively even if they are embedded in different departments with ranging academic cultures.

@@

Lord Kelvin's interest in the age of the Earth began in the early 1840s when he applied Fourier's theory of heat flow (i.e., heat flow by conduction) to various idealized systems. By 1844, he specifically applied his mathematics to Earth, assuming it to be a cooling conducting half-space with no internal heat flow. The mathematics did not yield physical solutions for an infinitely old Earth, which led Kelvin to conclude that Earth's age must be finite. Further, Kelvin argued, in his 1846 dissertation to Glasgow University, that he could calculate the age of the Earth with measurements of Earth's modern heat flow in the crust. However, at the time, heat flow measurements did not exist, so Kelvin's estimates for Earth's age based on these calculations had to wait.

@@

\begin{quote}
  For eighteen years it has pressed on my mind, that essential principles of Thermo-dynamics have been overlooked by those geologists who uncompromisingly oppose all paroxysmal hypotheses, and maintain not only that we have examples now before us, on the earth, of all the different actions by which its crust has been modified in geological history, but that these actions have never, or have not on the whole, been more violent in the past than they are at present.
\end{quote}
Here, Kelvin is criticizing the geologic paradigm of uniformitarianism, which was most famously outlined in Charles Lyell's book \emph{Principle of Geology} \citep{Lyell_1833}, and adopted by many geologists during this era. Below, the next few paragraphs take a short detour to explain Lyell's uniformitarianism because it is necessary to understand Kelvin's 1960s debate with geologist's over the age of the Earth.

@@

With this context, we can now we can return to Kelvin's 1862 thermodynamic calculations for a 20 to 400 million year old Earth. Throughout the mid-1860s, Kelvin's arguments appear to have been largely ignored by the geologic community. Part of the reason why geologists ignored Kelvin was because within the paradigm of Lyell's uniformitarianism, there was a belief that the Earth was best understood through qualitative rather than quantitative observations (e.g., see the previous quote from \emph{Principles of Geology}). Furthermore geologists lacked the mathematical skills to evaluate Kelvin's claims. This dynamic is illustrated by an interaction between Kelvin and the prominent geologist Andrew Ramsay at a 1867 conference in Dundee, Scotland. In response to Kelvin's calculations, Ramsay said, ``I am as incapable of estimating and understanding the reasons which you physicists have for limiting geological time as you are incapable of understanding the geological reasons for our unlimited estimates.''

Darwin was an exception, because he acknowledged Kelvin's arguments for a young Earth, and worried Kelvin's claims were incompatible with natural selection if the rate of evolution was slow. In a letter to Alfred Russel Wallace in 1969, Darwin wrote, ``Thomson's views of the recent age of the world have been for some time one of my sorest troubles'' \citep{Marchant_1916}. Darwin repeats his concern of Kelvin's work in the fifth and sixth editions of \emph{On the Origin of Species} \citep{Darwin_1869,Darwin_1872}. However, he also notes that biologists have only observed changes in species in the stratigraphic record, and have not quantified the rate of evolution measured by years. Therefore, according to Darwin, a young Earth with quick evolution is compatible with all the evidence. Furthermore, Darwin is generally skeptical of a physics-based calculation and wonders if Kelvin's calculation has bigger uncertainty than portrayed: ``...many philosophers are not as yet willing to admit that we know enough of the constitution of the universe and of the interior of our globe to speculate with safety on its past duration'' \citep{Darwin_1872}.

Through the late 1960s Kelvin published several more papers, repeating his 1962 arguments. Ultimately, the biologist Thomas Huxley used his presidential address to the London Geologic society to directly challenge Kelvin. He brought up Kelvin's changing theories for the Sun's heat and suggested that Kelvin's thermodynamic calculations of the cooling Earth had more uncertainty than Kelvin portrayed. Huxley also noted that no one believed strictly in Lyell's uniformitarianism and that geologists were generally comfortable with a 100-million-year-old Earth. However, this statement was not completely true because Darwin was clearly concerned based on his letters to Wallace. To conclude, Huxley stated:

@@

The debate over the age of the Earth was one of the earliest attempts of interdisciplinary science. The problem involved Kelvin's thermodynamic arguments, the geologic record, as well as Darwin's theory of biological evolution. However, this interdisciplinary experiment was not very successful. For years after his initial 1862 arguments for a young Earth, Kelvin was unable to convince geologists to take his calculations seriously. Also, Kelvin seemed to value his thermodynamics far more than evidence preserved in the stratigraphic record. Why couldn't any physicist, geologist or biologist consider all the evidence (geologic, or physics-based) to estimate Earth's age?

@@

A further problem is that specialization incentivizes researchers to emphasize or over-weight evidence from their domain of expertise. Consider the putative discovery of methane in Mars' atmosphere by the Curiosity rover mass spectrometer \citep{Webster_2015}. Methane on Mars is potentially a very important discovery, because on Earth, methane gas is associated with life. But the problem with the Curiosity observations is that the Mars Trace Gas Orbiter could not duplicate the observation \citep{Korablev_2019}. Also, methane at the concentrations suggested by Curiosity is not compatible with our understanding of methane photochemistry \citep{Lefevre_2009}. Despite this contradicting evidence researchers specialized in the Curiosity rover mass spectrometer are incentivized to champion the detection, because if there is indeed methane on Mars created by life, then the mass spectroscopists will be credited for the revolutionary discovery.

>>>>

Many of the big unsolved problems in science today will require an interdisciplinary approach to answer. For example, the unsolved problem of the origin of life on Earth can only be tackled using many different fields of research. Geologic expertise is required to understand the earliest fossil records of life, organic chemists are needed to synthesize life-like replicating molecules, and atmospheric modelers are essential to estimate the plausible conditions on the early Earth under which life began. Another problem that requires broad expertise is projecting the impact of anthropogenic climate change on society. Any forecast will involve atmospheric scientists, glaciologists, economists, and ecologists among many other areas of knowledge. Given that important unanswered problems rely on interdisciplinary evidence, it is essential that scientists work together effectively even if they are embedded in different departments with ranging academic cultures.

@@

Lord Kelvin's interest in the age of the Earth began in the early 1840s when he applied Fourier's theory of heat flow (i.e., heat flow by conduction) to various idealized systems. By 1844, he specifically applied his mathematics to Earth, assuming it to be a cooling conducting half-space with no internal heat flow. The mathematics did not yield physical solutions for an infinitely old Earth, which led Kelvin to conclude that Earth's age must be finite. Further, Kelvin argued, in his 1846 dissertation to Glasgow University, that he could calculate the age of the Earth with measurements of Earth's modern heat flow in the crust. However, at the time, heat flow measurements did not exist, so Kelvin's estimates for Earth's age on the basis of these calculations had to wait.

@@

\begin{quote}
  For eighteen years it has pressed on my mind, that essential principles of Thermo-dynamics have been overlooked by those geologists who uncompromisingly oppose all paroxysmal hypotheses, and maintain not only that we have examples now before us, on the earth, of all the different actions by which its crust has been modified in geological history, but that these actions have never, or have not on the whole, been more violent in the past than they are at present.
\end{quote}
Here, Kelvin is criticizing the geologic paradigm of uniformitarianism, which was most famously outlined in Charles Lyell's book \emph{Principle of Geology} \citep{Lyell_1833}, and adopted by many geologists during this era. The next few paragraphs explain Lyell's uniformitarianism because it is necessary to understand Kelvin's 1960s debate with geologist's over the age of the Earth.

@@

With this context, we can now we can return to Kelvin's 1862 thermodynamic calculations for a 20 to 400 million year old Earth. Throughout the mid-1860s, Kelvin's arguments appear to have been largely ignored by the geologic community. Part of the reason why geologists ignored Kelvin was because, within the paradigm of Lyell's uniformitarianism, mathematical calculations were not the best way to understand the Earth (e.g., see the previous quote from \emph{Principles of Geology}). Furthermore geologists lacked the quantitative skills to evaluate Kelvin's claims. An interaction between Kelvin and the prominent geologist Andrew Ramsay at a 1867 conference in Dundee illustrates this general sentiment. In response to Kelvin's calculations, Ramsay said, ``I am as incapable of estimating and understanding the reasons which you physicists have for limiting geological time as you are incapable of understanding the geological reasons for our unlimited estimates.''

Darwin was an exception among geologists, because he acknowledged Kelvin's arguments for a young Earth, and worried Kelvin's claims were incompatible with natural selection if the rate of evolution was slow. In a letter to Alfred Russel Wallace in 1969, Darwin wrote, ``Thomson's views of the recent age of the world have been for some time one of my sorest troubles'' \citep{Marchant_1916}. Darwin repeats his concern of Kelvin's work in the fifth and sixth editions of \emph{On the Origin of Species} \citep{Darwin_1869,Darwin_1872}. However, he also notes that biologists have only observed species change in the stratigraphic record, and have not quantified the rate of evolution measured by years. Therefore, according to Darwin, a young Earth with quick evolution is compatible with all the evidence. Furthermore, Darwin is generally skeptical of a physics-based calculation and wonders if Kelvin's calculation has bigger uncertainty than portrayed: ``...many philosophers are not as yet willing to admit that we know enough of the constitution of the universe and of the interior of our globe to speculate with safety on its past duration'' \citep{Darwin_1872}.

Through the late 1960s Kelvin published several more papers, repeating his 1962 arguments. Ultimately, the biologist Thomas Huxley used his presidential address to the London Geologic society to directly challenge Kelvin. He brought up Kelvin's changing theories for the Sun's heat and suggested that Kelvin's thermodynamic calculations of the cooling Earth had more uncertainty than Kelvin admitted. Huxley also noted that no one believed strictly in Lyell's uniformitarianism and that geologists were generally comfortable with a 100-million-year-old Earth. However, this statement was not completely true because Darwin was clearly concerned based on his letters to Wallace. To conclude, Huxley stated:

@@

The debate over the age of the Earth was one of the earliest attempts of interdisciplinary science. The problem involved Kelvin's thermodynamic arguments, the geologic record, as well as Darwin's theory of biological evolution. However, this interdisciplinary experiment was not very successful. For years after his initial 1862 arguments for a young Earth, Kelvin was unable to convince geologists to take his calculations seriously. Also, Kelvin seemed to value his thermodynamics far more than evidence preserved in the stratigraphic record. Why couldn't any physicist, geologist or biologist consider all the evidence to estimate Earth's age?

@@

A further problem is that specialization incentivizes researchers to emphasize or over-weight evidence from their domain of expertise. Consider the claimed discovery of methane in Mars' atmosphere by the Curiosity rover mass spectrometer \citep{Webster_2015}. Methane on Mars is potentially a very important discovery, because on Earth, methane gas is associated with life. But the problem with the Curiosity observations is that the Mars Trace Gas Orbiter could not duplicate the observation \citep{Korablev_2019}. Also, methane at the concentrations suggested by Curiosity is not compatible with our understanding of methane photochemistry \citep{Lefevre_2009}. The Curiosity rover mass spectroscopists are incentivized to champion their detection and ignore the contradicting evidence, because if there is indeed methane on Mars created by life, then the mass spectroscopists will be credited for a revolutionary discovery.
